Orienting the causal relationship between imprecisely measured traits using GWAS summary data

### 2017-10-24 docs/manuscript.tex
updated, preparing for proofs

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using GWAS summary data}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{10 October 2017}

@@

Observational measures of the human phenome are growing ever more
abundant, but using these data to make causal inference is notoriously
susceptible to many pitfalls, with basic regression-based techniques
unable to distinguish a true causal association from reverse causation
or confounding (1--3). In response to this, the use of genetic
associations to instrument traits has emerged as a technique for
improving the reliability of causal inference in observational data, and
with the coincident rise in genome-wide association studies it is now a
prominent tool that is applied in several different guises (3--6).
However, shifting from observational associations to instrumentation
does require more (often untestable) assumptions, and potential pitfalls
remain. One that is often neglected is the influence of non-differential
measurement error on the reliability of causal inference.

Measurement error is the difference between the measured value of a
quantity and its true value. This study focuses specifically on
non-differential measurement error where all strata of a measured
variable have the same error rate, which can manifest as changes in
scale or measurement imprecision (noise). Such variability can arise
through a whole plethora of mechanisms, which are often specific to the
study design and difficult to avoid (7,8). Array technology is now
commonly used to obtain high throughput phenotyping at low cost, but
comes with the problem of having imperfect resolution, for instance
methylation levels as measured by the Illumina450k chip are prone to
have some amount of noise around the true value due to imperfect
sensitivity (9,10). Relatedly, if the measurement of biological interest
is the methylation level in a T cell, then measurement error of this
value can be introduced by using methylation levels from whole blood
samples because the measured value will be an assay of many cell types
(11).

Measurement error will of course arise in other types of data too. For
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect (12), leading to the problem
that phenotypes may be imprecisely defined. A similar problem of
biological misspecification is unavoidable in disease diagnosis, and
measuring behaviour such as smoking or diet is notoriously difficult to
do accurately. Measurement error can also be introduced after the data
have been collected, for example the transformation of non-normal data
for the purpose of statistical analysis will lead to a new variable that
will typically incur both changes in scale and imprecision (noise)
compared to the original variable. The sources of measurement error are
not limited to this list (8), and its impact has been explored in the
epidemiological literature extensively (13,14). Given the
near-ubiquitous presence of measurement error in phenomic data it is
vital to understand its impact on the tools we use for causal inference.

An established study design that can provide information about causality
is randomisation. Given the hypothesis that trait A (henceforth referred
to as the exposure) is causally related to trait B (henceforth referred
to as the outcome), randomisation can be employed to assess the causal
nature of the association by randomly splitting the sample into two
groups, subjecting one group to the exposure and treating the other as a
control. The association between the exposure and the outcome in this
setting provides a robust estimate of the causal relationship. This
provides the theoretical basis behind randomised control trials, but in
practice randomisation is often difficult or impossible to implement in
an experimental context due to cost, scale or inability to manipulate
the exposure. The principle, however, can be employed in extant
observational data through the use of genetic variants associated with
the exposure (instruments), where the inheritance of an allele serves as
a random lifetime allocation of differential exposure levels (15,16).
Two statistical approaches to exploiting the properties of genetic
instruments are widely used: mediation-based approaches and Mendelian
randomisation (MR).

Mediation-based approaches employ genetic instruments (typically single
nucleotide polymorphisms, SNPs) to orient the causal direction between
the exposure and the outcome. If a SNP is associated with an exposure,
and the exposure is associated with some outcome, then it logically
follows that in this simple three-variable scenario the estimated direct
influence of the SNP on the outcome will be zero when conditioning on
the exposure. Here, the exposure completely mediates the association
between the SNP and the outcome, providing information about the causal
influence of the exposure on the outcome. This forms the basis of a
number of methods such as genetical genomics (17), the regression-based
causal inference test (CIT) (4,18), a structural equation modelling
(SEM) implementation in the NEO software (5), and various other methods
including Bayesian approaches (6). They have been employed by a number
of recent publications that make causal inferences in large scale `omics
datasets (6,19--23).

MR can be applied to the same data - phenotypic measures of the exposure
and the outcome variables and a genetic instrument for the exposure -
but the genetic instrument is employed in a subtly different manner.
Here the SNP is used as a surrogate for the exposure. Assuming the SNP
associates with the outcome only through the exposure, the causal effect
of the exposure on the outcome can be estimated by scaling the
association between the SNP and the outcome by the association between
the SNP and the exposure. Though difficult to test empirically, this
assumption can be relaxed in various ways when multiple instruments are
available for a putative exposure (24,25) and a number of sensitivity
tests are now available to improve reliability (26). Additionally, if
valid genetic instruments are known for both traits of interest then MR
can be performed in both directions (bi-directional MR), testing the
influence of one trait on the other and vice versa, to infer the causal
direction between the two phenotypes (27,28).

@@

Assuming biological knowledge of genetic associations can be problematic
because if there exists a putative association between two variables,
with the SNP being robustly associated with each, it can be difficult to
determine which of the two variables is subject to the primary effect of
the SNP (i.e.~for which of the two variables is the SNP a valid
instrument? Figure 1). By definition, we expect that if the association
is causal then a SNP for the exposure will be associated with the
outcome, such that if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Genome-wide
association studies (GWASs) that identify genetic associations for
complex traits are, by design, hypothesis free and agnostic of genomic
function, and it often takes years of follow up studies to understand
the biological nature of a putative GWAS hit (29). If multiple
instruments are available for an hypothesised exposure, which is
increasingly typical for complex traits that are analysed in large GWAS
consortia, then techniques can be applied to mitigate these issues (16).
But these techniques cannot always be applied in the case of determining
causal directions between 'omic measures where typically only one
cis-acting SNP is known. For example if a DNA methylation probe is
associated with expression of an adjacent gene, then is a cis-acting SNP
an instrument for the DNA methylation level, or the gene expression
level (Figure 1)?

MR has some important advantages over the mediation-based approaches.
First, the mediation-based approaches require that the exposure, outcome
and instrumental variables are all measured in the same data, whereas
recent extensions to MR circumvent this requirement, allowing causal
inference to be drawn when exposure variables and outcome variables are
measured in different samples (30). This has the crucial advantage of
improving statistical power by allowing analysis in much larger sample
sizes, and dramatically expands the breadth of possible phenotypic
relationships that can be evaluated (26). Second, the mediation-based
approach of adjusting the outcome for the exposure to nullify the
association between the SNP and the outcome is affected by unmeasured
confounding of the exposure and outcome. This is because adjusting the
outcome by the exposure induces a collider effect between the SNP and
outcome (31), and in order to fully abrogate this association one must
also adjust for all (hidden or otherwise) confounders. MR does not
suffer from this problem because it does not test for association
through adjustment. Third, when MR assumptions are satisfied the method
is robust to there being measurement error in the exposure variable
(32). Indeed instrumental variable (IV) analysis was in part initially
introduced as a correction for measurement error in the exposure (33),
whereas it has been noted that both classic mediation-based analyses
(13,14,34,35) and mediation-based methods that use instrumental
variables (36,37) are prone to be unreliable in its presence.

Using theory and simulations we show how non-differential measurement
error in phenotypes can lead to unreliable causal inference in the
mediation-based CIT method. Though we only examine the CIT method in
detail, we believe that attempting to adjust for mediating variables to
make causal inference is susceptible to problems, which can be
generalised to other mediation-based methods. We then present an
extension to MR that allows researchers to ascertain the causal
direction of an association even when the biology of the instruments are
not fully understood, and also a metric to evaluate the sensitivity of
the result of this extension to measurement error. Finally, to
demonstrate the potential impact of measurement error we apply this
method to infer the direction of causation between DNA methylation
levels and gene expression levels. Our analyses highlight that because
these different causal inference techniques have varying strengths and
weaknesses, triangulation of evidence from as many sources as possible
should be practiced in causal inference (38).

@@

Measurement error of an exposure can be modeled as a transformation of
the true value (\(x\)) that leads to the observed value, \(x_o = f(x)\).
For example, following Pierce and VanderWeele (32) we can define

@@

In this study we assume there is no measurement error in the SNP. Common
genetic variants are typically less susceptible to measurement error due
to strict quality control procedures prior to genome wide association
studies. Any non-differential measurement error that might be present
(either because the SNP is poorly typed or because the SNP is not in
complete linkage disequilibrium with the causal variant) will reduce
power in MR but will not incur bias (3,13,32). We also assume that
measurement error in the exposure and the outcome are uncorrelated.

@@

First we describe how the CIT method (4) is implemented in the R package
\emph{R/cit} (18). The methodology of the CIT is as follows. Assume an
exposure \(x\) is instrumented by a SNP \(g\), and the exposure \(x\)
causes an outcome \(y\), as described above. The following tests are
then performed:

@@

The term in the 4th test can be rewritten as a
\(cov(g, y|x) = cov(g, y - \hat{y})\) where
\(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the residual
of \(y\) after adjusting for \(x\), and \(x\) is assumed to mediate the
association between the SNP and the outcome. The condition in the 4th
test is formulated as an equivalence testing problem that is estimated
using simulations, comparing the estimate from the data against
empirically obtained estimates for simulated variables where the
independence model is true (full details are given in (4)). We note here
that this approach is liable to fail, even when there is a true causal
relationship, when confounders of the exposure and outcome are present,
as these will induce collider bias.

@@

\begin{itemize}
\tightlist
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then model 1 is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then model 2 is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then no evidence for a causal
  relationship
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then there is potentially
  confounding (Supplementary figure 1) and no call is made.
\end{itemize}

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value, though we stress that for real analyses it is
not good practice to rely on p-values for making causal inference, nor
is it reliable to depend on arbitrary significance thresholds (39).

@@

Two stage least squares (2SLS) is a commonly used technique for
performing MR when the exposure, outcome and instrument data are all
available in the same sample. A p-value for this test, \(p_{MR}\), was
obtained using the \(systemfit\) function in the R package
\(R/systemfit\) (40). Note that the value of \(p_{MR}\) is identical
when using the same genetic variant to instrument the influence of the
exposure \(x\) on the outcome \(y\), or erroneously, instrumenting the
outcome \(y\) on the exposure \(x\).

@@

To infer the direction of causality it is desirable to know which of the
variables, \(x\) or \(y\), is being directly influenced by the
instrument \(g\). This can be achieved by assessing which of the two
variables has the biggest absolute correlation with \(g\) (Supplementary
text 2), formalised by testing for a difference in the correlations
\(\rho_{gx}\) and \(\rho_{gy}\) using Steiger's Z-test for correlated
correlations within a population (41). It is calculated as

@@

Note that the same correlation test approach can be applied to a
two-sample MR setting. Two-sample MR refers to the case where the
SNP-exposure association and SNP-outcome association are calculated in
different samples (e.g.~from publicly available summary statistics
(26,30)). Here the Steiger test of two independent correlations can be
applied where.

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package \emph{R/psych} (42).

The Steiger test assumes that there is a causal relationship between the
two variables, and that the SNP is a valid instrument for one of them.
However it is liable to give incorrect causal directions under some
other circumstances. First, some levels of horizontal pleiotropy, where
the SNP influences the outcome through some pathway other than the
exposure, could induce problems because this is a means by which the
instrument is invalid. Second, some differential values of measurement
error between the exposure and the outcome could induce incorrect causal
directions, and this is explored in Supplementary text 2. Third, some
levels of unmeasured confounding between the exposure and the outcome
could induce the wrong causal direction, and this is explored in
Supplementary text 3.

@@

As we show in Supplementary text 2 it is possible to infer the bounds of
measurement error on \(x_o\) or \(y_o\) given known genetic
associations. The maximum measurement imprecision of \(x_o\) is
\(\rho_{g,x_o}\), because it is known that at least that much of the
variance has been explained in \(x_o\) by \(g\). The minimum is 0,
denoting perfectly measured trait values (the same logic applies to
\(y_o\)). It is possible to simulate what the inferred causal direction
would be for all values within these bounds.

To evaluate how reliable, \(R\), the inference of the causal direction
is to potential measurement error in \(x\) and \(y\) we need to predict
the values of \(\rho_{gy} - \rho_{gx}\) for those values of measurement
error. We offer two tools in which to do this. First, the user can
provide values of measurement error for \(x\) and \(y\) and obtain a
revised inference of the causal direction. Second, we integrate over the
entire range of \(\rho_{gy} - \rho_{gx}\) values for possible
measurement error values, assuming that any measurement error value is
equally likely. Across all possible values of measurement error in \(x\)
and \(y\) we find the volume that agrees with the inferred direction of
causality and the volume that disagrees with the inferred direction of
causality, and take the ratio of these two values. A ratio \(R=1\)
indicates that the inferred causal direction is highly sensitive to
measurement error, because equal weight of the measurement error
parameter space supporting both directions of causality. In general, the
\(R\) value denotes that the inferred direction of causality is \(R\)
times more likely to be the empirical result than the opposite
direction. Full details are provided in Supplementary text 2.

@@

where non-differential measurement error is represented by a noise
(measurement imprecision) term
\(\epsilon_{m*} \sim N(0, \sigma^2_{m*})\), and measurement bias terms
\(\alpha_{m*}\) and \(\beta_{m*}\) for the exposure variable \(x\) and
the outcome variable \(y\). Note that following the first section of the
Results we no longer include the bias terms for simplicity. We have
formulated the the non-causal model as:

@@

giving a total of 432 combinations of parameters. Simulations using each
of these sets of variables were performed 100 times, and the CIT and MR
methods were applied to each in order to evaluate the causal association
of the simulated variables. Similar patterns of results were obtained
for different values of \(cor(g, x)\) and \(cor(g, u)\).

@@

Two sample MR (30) was performed using the summary statistics for
genetic influences on gene expression and DNA methylation. To do this we
obtained a list of 458 gene expression - DNA methylation associations as
reported in Shakhbazov et al (43). These were filtered to be located on
the same chromosome, have robust correlations after correcting for
multiple testing, and to share a SNP that had a robust cis-acting effect
on both the DNA methylation probe and the gene expression probe. Because
only summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

All analysis was performed using the R programming language (44) and
code is made available at
\url{https://github.com/explodecomputer/causal-directions} and
implemented in the MR-Base (\url{http://wwww.mrbase.org}) platform (26).

@@

In the causal inference test (CIT), the 4th condition (see Methods)
employs mediation for causal inference, and can be expressed as
\(cov(g, y - \hat{y}) = 0\), where
\(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement error
in scale and imprecision is introduced, such that \(y_o\) is the
measured value of \(y\), it can be shown using basic covariance
properties (Supplementary text 1) that

@@

If we do not know whether the SNP \(g\) has a primary influence on \(x\)
or \(y\) then CIT can attempt to infer the causal direction. Though
bi-directional MR can be used to orient causal directions (27), this
requires knowledge of a valid instrument for each trait, and we were
motivated to develop the MR Steiger method that could operate on summary
data to orient the direction of causality using the same conditions as
the CIT, where the underlying biology of a single SNP is not fully
understood. We go on to explore the scenarios in which the method is
likely to return the correct or incorrect causal directions.

We performed simulations to compare the power and type 1 error rates of
MR and CIT in detecting a causal association between simulated variables
under different levels of imprecision simulated in the exposure.
Comparing the performance of methods with different sets of assumptions
can be difficult, but a basic comparison is shown in Figure 3. We
observe that the CIT is more conservative under the null model of no
association owing to the omnibus test statistic comprising several
statistical tests. The FDR using a p-value threshold of 0.05 appears to
be close to 0, whereas for the MR Steiger method the FDR is around 0.05.
Using the same p-value thresholds to declare significance in the
non-null simulations, the general trend appears to be that the CIT power
reduces as measurement error in the exposure increases more steeply than
that of the MR Steiger method.

For a particular association, it is of interest to identify the range of
possible measurement error values for which the method will give results
that agree or disagree with the empirically inferred causal direction
(Figure 4a, Supplementary text 2). This metric can be used to evaluate
the reliability of MR Steiger test.

>>>>

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using GWAS summary data}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{24 October 2017}

@@

Observational measures of the human phenome are growing ever more
abundant, but using these data to make causal inference is notoriously
susceptible to many pitfalls, with basic regression-based techniques
unable to distinguish a true causal association from reverse causation
or confounding {[}1--3{]}. In response to this, the use of genetic
associations to instrument traits has emerged as a technique for
improving the reliability of causal inference in observational data, and
with the coincident rise in genome-wide association studies it is now a
prominent tool that is applied in several different guises {[}3--6{]}.
However, shifting from observational associations to instrumentation
does require more (often untestable) assumptions, and potential pitfalls
remain. One that is often neglected is the influence of non-differential
measurement error on the reliability of causal inference.

Measurement error is the difference between the measured value of a
quantity and its true value. This study focuses specifically on
non-differential measurement error where all strata of a measured
variable have the same error rate, which can manifest as changes in
scale or measurement imprecision (noise). Such variability can arise
through a whole plethora of mechanisms, which are often specific to the
study design and difficult to avoid {[}7,8{]}. Array technology is now
commonly used to obtain high throughput phenotyping at low cost, but
comes with the problem of having imperfect resolution, for instance
methylation levels as measured by the Illumina450k chip are prone to
have some amount of noise around the true value due to imperfect
sensitivity {[}9,10{]}. Relatedly, if the measurement of biological
interest is the methylation level in a T cell, then measurement error of
this value can be introduced by using methylation levels from whole
blood samples because the measured value will be an assay of many cell
types {[}11{]}.

Measurement error will of course arise in other types of data too. For
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect {[}12{]}, leading to the problem
that phenotypes may be imprecisely defined. A similar problem of
biological misspecification is unavoidable in disease diagnosis, and
measuring behaviour such as smoking or diet is notoriously difficult to
do accurately. Measurement error can also be introduced after the data
have been collected, for example the transformation of non-normal data
for the purpose of statistical analysis will lead to a new variable that
will typically incur both changes in scale and imprecision (noise)
compared to the original variable. The sources of measurement error are
not limited to this list {[}8{]}, and its impact has been explored in
the epidemiological literature extensively {[}13,14{]}. Given the
near-ubiquitous presence of measurement error in phenomic data it is
vital to understand its impact on the tools we use for causal inference.

An established study design that can provide information about causality
is randomisation. Given the hypothesis that trait A (henceforth referred
to as the exposure) is causally related to trait B (henceforth referred
to as the outcome), randomisation can be employed to assess the causal
nature of the association by randomly splitting the sample into two
groups, subjecting one group to the exposure and treating the other as a
control. The association between the exposure and the outcome in this
setting provides a robust estimate of the causal relationship. This
provides the theoretical basis behind randomised control trials, but in
practice randomisation is often difficult or impossible to implement in
an experimental context due to cost, scale or inability to manipulate
the exposure. The principle, however, can be employed in extant
observational data through the use of genetic variants associated with
the exposure (instruments), where the inheritance of an allele serves as
a random lifetime allocation of differential exposure levels
{[}15,16{]}. Two statistical approaches to exploiting the properties of
genetic instruments are widely used: mediation-based approaches and
Mendelian randomisation (MR).

Mediation-based approaches employ genetic instruments (typically single
nucleotide polymorphisms, SNPs) to orient the causal direction between
the exposure and the outcome. If a SNP is associated with an exposure,
and the exposure is associated with some outcome, then it logically
follows that in this simple three-variable scenario the estimated direct
influence of the SNP on the outcome will be zero when conditioning on
the exposure. Here, the exposure completely mediates the association
between the SNP and the outcome, providing information about the causal
influence of the exposure on the outcome. This forms the basis of a
number of methods such as genetical genomics {[}17{]}, the
regression-based causal inference test (CIT) {[}4,18{]}, a structural
equation modelling (SEM) implementation in the NEO software {[}5{]}, and
various other methods including Bayesian approaches {[}6{]}. They have
been employed by a number of recent publications that make causal
inferences in large scale `omics datasets {[}6,19--23{]}.

MR can be applied to the same data - phenotypic measures of the exposure
and the outcome variables and a genetic instrument for the exposure -
but the genetic instrument is employed in a subtly different manner.
Here the SNP is used as a surrogate for the exposure. Assuming the SNP
associates with the outcome only through the exposure, the causal effect
of the exposure on the outcome can be estimated by scaling the
association between the SNP and the outcome by the association between
the SNP and the exposure. Though difficult to test empirically, this
assumption can be relaxed in various ways when multiple instruments are
available for a putative exposure {[}24,25{]} and a number of
sensitivity tests are now available to improve reliability {[}26{]}.
Additionally, if valid genetic instruments are known for both traits of
interest then MR can be performed in both directions (bi-directional
MR), testing the influence of one trait on the other and vice versa, to
infer the causal direction between the two phenotypes {[}27,28{]}.

@@

Assuming biological knowledge of genetic associations can be problematic
because if there exists a putative association between two variables,
with the SNP being robustly associated with each, it can be difficult to
determine which of the two variables is subject to the primary effect of
the SNP (i.e.~for which of the two variables is the SNP a valid
instrument? Figure 1). By definition, we expect that if the association
is causal then a SNP for the exposure will be associated with the
outcome, such that if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Genome-wide
association studies (GWASs) that identify genetic associations for
complex traits are, by design, hypothesis free and agnostic of genomic
function, and it often takes years of follow up studies to understand
the biological nature of a putative GWAS hit {[}29{]}. If multiple
instruments are available for an hypothesised exposure, which is
increasingly typical for complex traits that are analysed in large GWAS
consortia, then techniques can be applied to mitigate these issues
{[}16{]}. But these techniques cannot always be applied in the case of
determining causal directions between 'omic measures where typically
only one cis-acting SNP is known. For example if a DNA methylation probe
is associated with expression of an adjacent gene, then is a cis-acting
SNP an instrument for the DNA methylation level, or the gene expression
level (Figure 1)?

MR has some important advantages over the mediation-based approaches.
First, the mediation-based approaches require that the exposure, outcome
and instrumental variables are all measured in the same data, whereas
recent extensions to MR circumvent this requirement, allowing causal
inference to be drawn when exposure variables and outcome variables are
measured in different samples {[}30{]}. This has the crucial advantage
of improving statistical power by allowing analysis in much larger
sample sizes, and dramatically expands the breadth of possible
phenotypic relationships that can be evaluated {[}26{]}. Second, the
mediation-based approach of adjusting the outcome for the exposure to
nullify the association between the SNP and the outcome is affected by
unmeasured confounding of the exposure and outcome. This is because
adjusting the outcome by the exposure induces a collider effect between
the SNP and outcome {[}31{]}, and in order to fully abrogate this
association one must also adjust for all (hidden or otherwise)
confounders. MR does not suffer from this problem because it does not
test for association through adjustment. Third, when MR assumptions are
satisfied the method is robust to there being measurement error in the
exposure variable {[}32{]}. Indeed instrumental variable (IV) analysis
was in part initially introduced as a correction for measurement error
in the exposure {[}33{]}, whereas it has been noted that both classic
mediation-based analyses {[}13,14,34,35{]} and mediation-based methods
that use instrumental variables {[}36,37{]} are prone to be unreliable
in its presence.

Using theory and simulations we show how non-differential measurement
error in phenotypes can lead to unreliable causal inference in the
mediation-based CIT method. Though we only examine the CIT method in
detail, we believe that attempting to adjust for mediating variables to
make causal inference is susceptible to problems, which can be
generalised to other mediation-based methods. We then present an
extension to MR that allows researchers to ascertain the causal
direction of an association even when the biology of the instruments are
not fully understood, and also a metric to evaluate the sensitivity of
the result of this extension to measurement error. Finally, to
demonstrate the potential impact of measurement error we apply this
method to infer the direction of causation between DNA methylation
levels and gene expression levels. Our analyses highlight that because
these different causal inference techniques have varying strengths and
weaknesses, triangulation of evidence from as many sources as possible
should be practiced in causal inference {[}38{]}.

@@

Measurement error of an exposure can be modeled as a transformation of
the true value (\(x\)) that leads to the observed value, \(x_o = f(x)\).
For example, following Pierce and VanderWeele {[}32{]} we can define

@@

In this study we assume there is no measurement error in the SNP. Common
genetic variants are typically less susceptible to measurement error due
to strict quality control procedures prior to genome wide association
studies. Any non-differential measurement error that might be present
(either because the SNP is poorly typed or because the SNP is not in
complete linkage disequilibrium with the causal variant) will reduce
power in MR but will not incur bias {[}3,13,32{]}. We also assume that
measurement error in the exposure and the outcome are uncorrelated.

@@

First we describe how the CIT method {[}4{]} is implemented in the R
package \emph{R/cit} {[}18{]}. The methodology of the CIT is as follows.
Assume an exposure \(x\) is instrumented by a SNP \(g\), and the
exposure \(x\) causes an outcome \(y\), as described above. The
following tests are then performed:

@@

The term in the 4th test can be rewritten as a
\(cov(g, y|x) = cov(g, y - \hat{y})\) where
\(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the residual
of \(y\) after adjusting for \(x\), and \(x\) is assumed to mediate the
association between the SNP and the outcome. The condition in the 4th
test is formulated as an equivalence testing problem that is estimated
using simulations, comparing the estimate from the data against
empirically obtained estimates for simulated variables where the
independence model is true (full details are given in {[}4{]}). We note
here that this approach is liable to fail, even when there is a true
causal relationship, when confounders of the exposure and outcome are
present, as these will induce collider bias.

@@

\begin{itemize}
\tightlist
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then model 1 is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then model 2 is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then no evidence for a causal
  relationship
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then there is potentially
  confounding (S1 fig) and no call is made.
\end{itemize}

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value, though we stress that for real analyses it is
not good practice to rely on p-values for making causal inference, nor
is it reliable to depend on arbitrary significance thresholds {[}39{]}.

@@

Two stage least squares (2SLS) is a commonly used technique for
performing MR when the exposure, outcome and instrument data are all
available in the same sample. A p-value for this test, \(p_{MR}\), was
obtained using the \(systemfit\) function in the R package
\(R/systemfit\) {[}40{]}. Note that the value of \(p_{MR}\) is identical
when using the same genetic variant to instrument the influence of the
exposure \(x\) on the outcome \(y\), or erroneously, instrumenting the
outcome \(y\) on the exposure \(x\).

@@

To infer the direction of causality it is desirable to know which of the
variables, \(x\) or \(y\), is being directly influenced by the
instrument \(g\). This can be achieved by assessing which of the two
variables has the biggest absolute correlation with \(g\) (S2 Text),
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population {[}41{]}. It is calculated as

@@

Note that the same correlation test approach can be applied to a
two-sample MR setting. Two-sample MR refers to the case where the
SNP-exposure association and SNP-outcome association are calculated in
different samples (e.g.~from publicly available summary statistics
{[}26,30{]}). Here the Steiger test of two independent correlations can
be applied where.

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package \emph{R/psych} {[}42{]}.

The Steiger test assumes that there is a causal relationship between the
two variables, and that the SNP is a valid instrument for one of them.
However it is liable to give incorrect causal directions under some
other circumstances. First, some levels of horizontal pleiotropy, where
the SNP influences the outcome through some pathway other than the
exposure, could induce problems because this is a means by which the
instrument is invalid. Second, some differential values of measurement
error between the exposure and the outcome could lead to incorrect
inference of the causal direction (S2 Text). Third, some levels of
unmeasured confounding between the exposure and the outcome could lead
to inference of the wrong causal direction (S3 Text).

@@

As we show in S2 Text it is possible to infer the bounds of measurement
error on \(x_o\) or \(y_o\) given known genetic associations. The
maximum measurement imprecision of \(x_o\) is \(\rho_{g,x_o}\), because
it is known that at least that much of the variance has been explained
in \(x_o\) by \(g\). The minimum is 0, denoting perfectly measured trait
values (the same logic applies to \(y_o\)). It is possible to simulate
what the inferred causal direction would be for all values within these
bounds.

To evaluate how reliable, \(R\), the inference of the causal direction
is to potential measurement error in \(x\) and \(y\) we need to predict
the values of \(\rho_{gy} - \rho_{gx}\) for those values of measurement
error. We offer two tools in which to do this. First, the user can
provide values of measurement error for \(x\) and \(y\) and obtain a
revised inference of the causal direction. Second, we integrate over the
entire range of \(\rho_{gy} - \rho_{gx}\) values for possible
measurement error values, assuming that any measurement error value is
equally likely. Across all possible values of measurement error in \(x\)
and \(y\) we find the volume that agrees with the inferred direction of
causality and the volume that disagrees with the inferred direction of
causality, and take the ratio of these two values. A ratio \(R=1\)
indicates that the inferred causal direction is highly sensitive to
measurement error, because equal weight of the measurement error
parameter space supports each direction of causality. In general, the
\(R\) value denotes that the inferred direction of causality is \(R\)
times more likely to be the empirical result than the opposite direction
(S2 Text).

@@

where non-differential measurement error is represented by a noise
(measurement imprecision) term
\(\epsilon_{m*} \sim N(0, \sigma^2_{m*})\), and measurement bias terms
\(\alpha_{m*}\) and \(\beta_{m*}\) for the exposure variable \(x\) and
the outcome variable \(y\). Note that following the first section of the
Results we no longer include the bias terms for simplicity. We have
formulated the non-causal model as:

@@

giving a total of 432 combinations of parameters. Simulations using each
of these sets of variables were performed 100 times, and the CIT and MR
methods were applied to each in order to evaluate the causal association
of the simulated variables. Similar patterns of results were obtained
for different values of \(cor(g, x)\).

@@

Two sample MR {[}30{]} was performed using summary statistics for
genetic influences on gene expression and DNA methylation. To do this we
obtained a list of 458 gene expression - DNA methylation associations as
reported in Shakhbazov et al {[}43{]}. These were filtered to be located
on the same chromosome, have robust correlations after correcting for
multiple testing, and to share a SNP that had a robust cis-acting effect
on both the DNA methylation probe and the gene expression probe. Because
only summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

All analysis was performed using the R programming language {[}44{]} and
code is made available at
\url{https://github.com/explodecomputer/causal-directions} and
implemented in the MR-Base (\url{http://wwww.mrbase.org}) platform
{[}26{]}.

@@

In the causal inference test (CIT), the 4th condition (see Methods)
employs mediation for causal inference, and can be expressed as
\(cov(g, y - \hat{y}) = 0\), where
\(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement error
in scale and imprecision is introduced, such that \(y_o\) is the
measured value of \(y\), it can be shown using basic covariance
properties (S1 Text) that

@@

If we do not know whether the SNP \(g\) has a primary influence on \(x\)
or \(y\) then CIT can attempt to infer the causal direction. Though
bi-directional MR can be used to orient causal directions {[}27{]}, this
requires knowledge of a valid instrument for each trait, and we were
motivated to develop the MR Steiger method that could operate on summary
data to orient the direction of causality using the same conditions as
the CIT, where the underlying biology of a single SNP is not fully
understood. We go on to explore the scenarios in which the method is
likely to return the correct or incorrect causal directions.

We performed simulations to compare the power and type 1 error rates of
MR and CIT in detecting a causal association between simulated variables
under different levels of imprecision simulated in the exposure.
Comparing the performance of methods with different sets of assumptions
can be difficult, but a basic comparison is shown in Figure 3. We
observe that the CIT is more conservative under the null model of no
association owing to the omnibus test statistic comprising several
statistical tests. The FDR using a p-value threshold of 0.05 appears to
be close to zero, whereas for the MR Steiger method the FDR is around
0.05. Using the same p-value thresholds to declare significance in the
non-null simulations, the general trend appears to be that the CIT power
reduces as measurement error in the exposure increases more steeply than
that of the MR Steiger method.

For a particular association, it is of interest to identify the range of
possible measurement error values for which the method will give results
that agree or disagree with the empirically inferred causal direction
(Figure 4a, S2 Text). This metric can be used to evaluate the
reliability of MR Steiger test.

We show that in the presence of measurement imprecision,
\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (S2 Text) determines the
range of parameters around which the MR Steiger test is liable to
provide the wrong direction of causality (\emph{i.e.} if \(d>0\) then
the MR Steiger test is likely to be correct about the causal direction).
Figure 4b shows that when there is no measurement error in \(x\), the MR
Steiger test is unlikely to infer the wrong direction of causality even
if there is measurement error in \(y\). It also shows that in most cases
where \(x\) is measured with error, especially when the causal effect
between \(x\) and \(y\) is not very large, the sensitivity of the MR
Steiger test to measurement error is relatively low.

Unmeasured confounding between the exposure and outcome can also give
rise to problems with the MR Steiger approach (S3 Text). The
relationship between unmeasured confounding and causal orientation is
complex across the parameter space of possible confounding values (S2
Fig). Based on the range of parameter values that we explored, when the
magnitude of the observational variance explained between the exposure
and the outcome is below 0.2 the MR Steiger method is unlikely to return
the incorrect causal direction due to unmeasured confounding.
